When Subgraph Isomorphism is Really Hard, and Why This Matters for Graph Databases

### 2018-03-07 paper/paper.tex
Formatting comments

\jairheading{1}{2017}{0-0}{0/0}{0/0}
\ShortHeadings{When Subgraph Isomorphism is Really Hard}
{McCreesh, Prosser, Solnon \& Trimble}
\firstpageno{1}

@@

\newcommand*\samethanks[1][\value{footnote}]{\footnotemark[#1]}

@@

\author{\name Ciaran McCreesh \thanks{This work was supported by the Engineering and Physical Sciences
           Research Council [grant numbers EP/K503058/1, EP/M508056/1, and EP/P026842/1]} \email
           ciaran.mccreesh@glasgow.ac.uk \\
       \addr University of Glasgow, Glasgow, Scotland \\
       \name Patrick Prosser \samethanks[1] \email patrick.prosser@glasgow.ac.uk \\
       \addr University of Glasgow, Glasgow, Scotland \\
       \name Christine Solnon \thanks{This work was supported
       by the ANR project SoLStiCe (ANR-13-BS02-0002-01)} \email christine.solnon@insa-lyon.fr \\
       \addr INSA-Lyon, LIRIS, UMR5205, F-69621, France \\
       \name James Trimble \samethanks[1] \email j.trimble.1@research.gla.ac.uk \\
       \addr University of Glasgow, Glasgow, Scotland}
\maketitle

@@

\bigskip

Parts of this paper were presented at IJCAI \cite{DBLP:conf/ijcai/McCreeshPT16}; the first three
sections have been expanded, whilst
\cref{section:regular,section:labelled,section:filterverify} are new material. Most of this work
also appeared in a PhD thesis \cite{o:McCreesh17}.

@@

\section{Non-Induced Subgraph Isomorphisms}\label{section:non-induced}

@@

The line plots arithmetic mean search effort using the Glasgow solver: for sparse patterns, the problem is
trivial; for dense patterns, proving unsatisfiability is not particularly difficult; and we see a
complexity peak around the point where half the instances are satisfiable.  We also plot the search
cost of individual instances, as points. The behaviour we observe looks remarkably similar to random
3SAT problems---compare, for example, Figure~1 of \citeA{DBLP:journals/cacm/Leyton-BrownHHX14}. In
particular, satisfiable instances tend to be easier, but show greater variation than unsatisfiable
instances, and there are exceptionally hard satisfiable instances \cite{DBLP:conf/cp/SmithG97}.

@@

For larger patterns, more complex behaviour emerges. If we are in the intersection of the bottom half
and the bottom right diagonal of the search space, we should always retain the original heuristic,
and if we are in the intersection of the top half and the top left diagonal, we should always use
the complements. This behaviour can be predicted by taking the partial derivatives of $\langle Sol
\rangle$ in the $-d_p$ and $d_t$ directions.  However, when inside the remaining two eighths of the
parameter space, the partial derivatives of $\langle Sol \rangle$ disagree on which heuristic to
use, and using directional derivatives is not enough to resolve the problem. A close observation of
the data suggests that the actual location of the phase transition may be involved (and perhaps
\citeauthor{DBLP:conf/aaai/Walsh98}'s suggestion applies only in these conditions). In any case, $\langle Sol
\rangle$ is insufficient to explain the observed behaviour in these two eighths of the parameter space.

@@

The \emph{association graph encoding} of a subgraph isomorphism problem is a reduction to the clique
decision problem. \citeA{DBLP:conf/cp/McCreeshNPS16} describe and study this approach in more
detail. Briefly, the association graph is constructed by creating a new graph with a vertex for each
pair $(p, t)$ of vertices from the pattern and target graphs respectively. There is an edge between
vertex $(p_1, t_1)$ and vertex $(p_2, t_2)$ if mapping $p_1$ to $t_1$ and $p_2$ to $t_2$
simultaneously is permitted, i.e.\ $p_1$ is adjacent to $p_2$ if and only if $t_1$ is adjacent to
$t_2$. A clique of size equal to the order of the pattern graph exists in the association graph if
and only if the problem is satisfiable \cite{o:Levi73}. We used this encoding with an
implementation of \citeA{DBLP:journals/cor/SegundoRJ11}'s bit-parallel maximum clique algorithm
BBMC, modified to solve the decision problem rather than the optimisation problem.  Again, our
results show that the instances in the central region remain hard, and additionally, some of the
easy unsatisfiable instances become hard.

@@

These experiments further highlight that VF2 occasionally finds some instances which should be easy hard.
But does this cause problems in practice? The literature suggests that it does. For example,
\citeA{o:Gromping14} uses VF2 in a package for the R statistics language, and states
\begin{displayquote}``There are (not so many) instances for which creation of a clear design is
    prohibitively slow in the current implementation that evaluates subgraph isomorphism with the
    VF2 algorithm \ldots Recent experiences with a few of these showed that the LAD algorithm was
    very fast in ruling out impossible matches, where VF2 took a long time.''\end{displayquote}
Similarly, \citeA{o:Murray12} uses VF2 inside a compiler, and observes extremely variable
(and prohibitively high) compile times in some cases due to the expense of subgraph isomorphism
calls---given the heavily labelled nature of these graphs, we conjecture that any domain-based
algorithm would eliminate this cost.

@@

Subsequently, in a widely cited\footnote{685 citations according to Google Scholar in February
2018.} paper, \citeA{DBLP:conf/sigmod/YanYH04} introduce an indexing system
called gIndex. Again, this system handles queries by first producing a set of candidates by
eliminating certain unsatisfiable instances, this time by using substructures. They argue that the
query response time, which is to be minimised, is governed by the equation \[ T =
T_{\mathit{search}} + \left|C_q\right| \cdot T_{\mathit{iso\_test}} \textnormal{,}\] where
$T_{\mathit{search}}$ is the time taken to search for a candidate set of potential solutions of size
$\left|C_q\right|$, and $T_{\mathit{iso\_test}}$ is the cost of a subgraph isomorphism test. They
reason that since subisomorphism testing is \NP-complete, by making $\left|C_q\right|$ as small as
possible, the query response time will be reduced.  (Importantly, it is \emph{not} the time taken to
load graphs from disk which contributes to the per-candidate cost, but rather the time to perform a
subgraph isomorphism call.) They conclude that ``graph indexing plays a critical role at efficient
query processing in graph databases''.

This equation is repeated and expanded upon by \citeA{DBLP:journals/tods/YanYH05} to explicitly
separate I/O and subisomorphism testing costs into $T_{\mathit{io}}$ and $T_{\mathit{iso\_test}}$
respectively, obtaining a query response time of \[
    T_{\mathit{search}} + \left|C_q\right| \cdot (T_{\mathit{io}} + T_{\mathit{iso\_test}})
    \textnormal{.} \] The authors explicitly state that ``the value of $T_{\mathit{iso\_test}}$ does
not change much for a given query''. The argument presented is that
\begin{displayquote}``Sequential scan is very costly
because one has to not only access the whole graph database but also check subgraph isomorphism. It
is known that subgraph isomorphism is an \NP-complete problem. Clearly, it is necessary to build
graph indices in order to help processing graph queries.''\end{displayquote}
We reassess these claims in light of
what we now know about the behaviour of modern subgraph isomorphism algorithms and the nature
of solving \NP-complete problems. Contrary to the authors' claims,
we do not expect $T_{\mathit{iso\_test}}$ to be anything like a constant, even if the orders of the
input graphs are similar. In particular, any instance which can be excluded based upon filtering
must have a very small proof of unsatisfiability. These instances \emph{should} be trivial with any
domain-based subgraph isomorphism algorithm. Thus, all filtering \emph{should} be doing is eliminating the startup
costs of a trivial subgraph isomorphism call. The fact that filtering was successful empirically
should make us wonder whether the subgraph isomorphism algorithm used was substantially
weaker than the state of the art.
Indeed, the algorithm is described only as ``the simplest approach'' in the
paper.  Additionally, the experiments focus on reducing the size of the candidate set, without
considering the time taken to verify different candidate set instances.  The claim that
$T_{\mathit{iso\_test}}$ does not change much is not justified experimentally, and no consideration
is given as to whether this would hold true for other subgraph isomorphism algorithms.

@@

The equation is also used by \citeA{DBLP:conf/icde/JiangWYZ07}, who claim that ``usually the
verification time dominates the Query Response Time [s]ince the computational complexity of
$T_{\mathit{iso\_test}}$ is \NP-Complete''. They note that \begin{displayquote}``Approximately, the
    value of $T_{\mathit{iso\_test}}$ does not change too much with the difference of query. Thus,
the key to reducing query response time is to minimize the size of the candidate answer
set''.\end{displayquote} This claim becomes understandable when one examines the subgraph
isomorphism algorithm chosen for the verification step \cite{DBLP:journals/jacm/Ullmann76}: as the
algorithm predates techniques like generalised arc-consistent all-different propagation
\cite{DBLP:conf/aaai/Regin94}, it cannot immediately detect unsatisfiability in simple cases like
there being two red vertices in the pattern but only one in the target.\footnote{Although
interestingly, this algorithm effectively maintains (binary) arc consistency, and has a
variable-ordering heuristic, before these concepts appeared in the constraints literature.}

Without using the equation, \citeA{DBLP:conf/sigmod/ChengKNL07} state that since subgraph
isomorphism is \NP-complete, processing by a sequential scan is infeasible. They introduce new
filtering techniques to try to avoid the subgraph isomorphism step. A similar claim is made by
\citeA{DBLP:conf/icde/ZhangHY07} in an introduction of another indexing technique: ``obviously it is
inefficient to perform a sequential scan on every graph in the database, because the subgraph
isomorphism test is expensive''.

Muddying the waters slightly, a survey by \citeA{DBLP:journals/datamine/HanCXY07} states that
``large volumes of data'' (not \NP-completeness) is the reason for using indexing in these systems.
However, in a description of a system tailored to biological networks,
\citeA{DBLP:conf/edbt/ZhangLY09} state that \begin{displayquote}``Since the size of the raw database
    graph is small, it can be easily fit in the main memory.  However, the query (matching) time
will be very long due to the \NP-hard complexity.''\end{displayquote} They suggest that indexing is
a way of avoiding this. Similarly, \citeA{DBLP:journals/pvldb/ZhaoH10} argue that ``the graph query
problem is hard in that \ldots it requires subgraph isomorphism checking \ldots which has proven to
be \NP-complete'' and that working with large networks is hard or impossible ``due to the lack of
scalable graph indexing mechanisms''.

\citeA{DBLP:conf/icdcs/CaoYWRL11} propose a privacy-preserving cloud graph database system using
filter~/ verify, stating that ``checking subgraph isomorphism is \NP-complete, and therefore it is
infeasible to employ such a costly solution'' which simply checks every graph for a match.
\citeA{DBLP:journals/tkde/WangWYY12} look at indexing large sparse graphs. They state that ``because
subgraph isomorphism is an \NP-complete problem, a filter-and-verification method is usually
employed to speed up the search efficiency of graph similarity matching over a graph set''.
Similarly, after reviewing the literature, \citeA{DBLP:journals/vldb/YuanM13} argue that subgraph
querying is costly because it is \NP-complete, and that indices can improve the performance of graph
database queries. Again, new indexing techniques are introduced. Subsequently,
\citeA{DBLP:journals/pvldb/KatsarouNT15} perform a comprehensive comparison of graph database
filtering techniques. They state that performing a query against each graph in the dataset
``obviously does not scale, as subgraph isomorphism is \NP-complete''. In describing a system which
returns a special subset of graphs which match a query, \citeA{DBLP:journals/tkde/ZhengLZHZ16}
suggest that it is ``\NP-hard to check the graph isomorphism'' (meaning subgraph isomorphism), and
``in order to improve the time efficiency'' they use an indexing system to ``avoid as many costly
subgraph isomorphism checkings as possible''. Their index takes tens of thousands of seconds to
build, and they suggest that Ullmann's algorithm and VF2 are state of the art for verification.
\citeA{DBLP:journals/www/PengZ0LZ16} state that ``obviously, it is impossible to employ some
subgraph isomorphism algorithm, such as Ullmann or VF2'', and argue that ``in order to speed up
query processing'', they need to create indices.  And \citeA{DBLP:conf/isda/AzaouziR16} argue that
``since the subgraph isomorphism test is expensive, checking all graphs of a large database can be
unfeasible'', saying that ``naturally, the verification step is computationally more expensive since
it requires a subgraph isomorphism''; their experiments look at candidate set reduction sizes, and
their choice of subgraph isomorphism algorithm is not mentioned.

The filter~/ verify paradigm also influences other research. For example, it is used by
\citeA{DBLP:conf/icde/TianP08} in an approximate subgraph searching system: they state that
Ullmann's algorithm ``is prohibitively expensive for querying against [a] database with a large
number of graphs'', and that indices are used ``to filter out graphs that do not match the query''.
More recently, \citeA{DBLP:journals/pvldb/YuanMG13} continue a line of supergraph search work, again
using a filtering step to avoid subgraph isomorphism calls (in the opposite direction, so queries
are now target graphs). \citeA{DBLP:journals/tkde/HongZLY15} look at graph database subgraph
isomorphism with an additional set-similarity constraint, and state that Ullmann's algorithm and VF2
are ``usually costly for large graphs'' because they ``do not utilize any index structure''. They
propose an indexing structure, which takes over 2,000 seconds to construct, and uses nearly 2GBytes
of space.  There is also research into maintaining indices when the set of target graph changes: for
example \citeA{DBLP:conf/sigmod/YuanMYG15} look at algorithms for updating graph indices.
And describing a system for reusing results of queries which are sub- or super-graphs of previous
queries, \citeA{DBLP:conf/edbt/WangNT16} state that querying is a ``very costly operation as it
entails the \NP-complete problem of subgraph isomorphism'', and place ``an emphasis on the number of
unnecessary subgraph isomorphism tests''.

@@

We saw in the previous sections that although permuting input graphs could improve VF2's behaviour
on some instances, doing so does not make its performance come close to that of domain-based
algorithms.  \citeauthor{DBLP:conf/edbt/KatsarouNT17} say they ``hope that our findings will open up
new research directions, striving to find appropriate, per-query, isomorphic rewritings, in
combination with alternate per-query sub-iso algorithms that can yield large improvements''. We
believe it is important instead to emphasise previous research directions that have already solved
most of this problem, and to help ensure that these techniques become more widely known.

\citeA{DBLP:conf/edbt/KatsarouNT17}'s second claim is that ``different
algorithms find different queries hard''. To address this, they run many subgraph isomorphism
algorithms and input permutations in parallel; the extensive literature on parallel portfolios in
general \cite{DBLP:journals/ai/GomesS01}, and the approach by \citeA{DBLP:conf/sls/BattitiM07} for
subgraph isomorphism in particular, is not noted. (Nor is the portfolios literature referenced when
they say that ``using machine learning models to predict which version \ldots to employ per query is
of high interest'').  The evidence so far in this paper suggests that we should carefully examine
the reasons behind these speedups. It is certainly true that there are some instances that all
algorithms find hard, and there are good theoretical reasons to believe that these instances
genuinely are really hard.  Furthermore, algorithm portfolios are also well-known to be a successful
technique, and are beneficial even with modern subgraph isomorphism algorithms
\cite{DBLP:conf/lion/KotthoffMS16}.  However, in \cref{section:labelled} we saw that there are many
instances that VF2 finds hard that should not be hard, and that are not hard for other algorithms.
If permuting graphs sometimes addresses the difficulty of some of these instances, then it is likely
that they are not genuinely hard instances at all. Perhaps then, it would be better simply to use an
algorithm with domain tracking and domain-based ordering heuristics.

To test this suggestion, we return briefly to the experiments on the datasets discussed earlier in
this section. What if we modify our solver so that it does not use ``smallest domain first'', and
does not detect domain wipeouts until attempting to branch on an empty variable?  In this case, three of
the four datasets become extremely hard, with many instances now timing out after making hundreds of
millions of recursive calls, and the PPI dataset now requires hundreds of thousands of calls for
some queries.  Furthermore, slight changes to the input vertex ordering can now make some of these
``hard'' instances easy again.  This verifies our intuition: the apparent success of
\citeA{DBLP:conf/edbt/KatsarouNT17}'s technique is due to the ease of sometimes getting better
results out of a poor algorithm. Obtaining huge improvements from an algorithm portfolio consisting
only of variations of poor algorithms is not sufficient to demonstrate a genuine improvement over
the state of the art.

@@

\begin{displayquote}
    ``Benchmark problems should be hard. I report on the solution of the five open benchmark
    problems introduced \ldots for testing bin packing problems. Since the solutions were found
    either by hand or by using very simple heuristic methods, these problems would appear to be
    easy. In four cases I give improved packings to refute conjectures that previously reported
    packings were optimal, and I give a proof that the fifth conjecture was correct. \ldots Future
    experimenters should be careful to perform tests on problems that can reasonably be regarded as
    hard.''
\end{displayquote}

@@

\citeA{o:CarlettiFSV17} correctly observe that ``a performance improvement on [hard instances] has a
far greater practical impact than it would have on easier graphs, since the saved time can be of
several hours or even days''. To evaluate VF3, they introduce a new dataset consisting entirely of
satisfiable instances created by taking a connected subgraph of a randomly generated graph and
permuting the nodes. They conclude that VF3 is the best algorithm based upon its performance solving
the enumeration problem on these new instances.  We urge a more cautious interpretation of these
results: as we saw in \cref{section:induced}, being genuinely hard (rather than simply larger or
denser) is a subtle property, and care must be taken when empirically evaluating algorithms. In
particular, if one only considers large, easy instances, then the simpler the algorithm, the better
it scales. Indeed, we saw in \cref{figure:induced} that VF3 performs \emph{worse} than VF2 on
genuinely harder instances, and additionally behaves erratically in parts of the parameter space
that every other solver finds easy; finally, an induced analogue of \cref{figure:labels} confirms
that VF3 also still struggles with many obviously unsatisfiable instances.

@@

\acks{The authors would like to thank Kitty Meeks and Craig Reilly for their comments. We are
grateful to Iva Babukova for her help with the datasets and for sharing the results of her
experiments. The heatmap schemes used in this paper were created with the assistance of the
``Chroma.js Color Scale Helper'' by Gregor Aisch.}

>>>>

\jairheading{61}{2018}{0-0}{11/17}{03/18}
\ShortHeadings{When Subgraph Isomorphism is Really Hard}
{McCreesh, Prosser, Solnon \& Trimble}
\firstpageno{1}

@@

@@

\author{\name Ciaran McCreesh \email ciaran.mccreesh@glasgow.ac.uk \\
       \name Patrick Prosser \email patrick.prosser@glasgow.ac.uk \\
       \addr University of Glasgow, Glasgow, Scotland
       \AND
       \name Christine Solnon \email christine.solnon@insa-lyon.fr \\
       \addr INSA-Lyon, LIRIS, UMR5205, F-69621, France
       \AND
       \name James Trimble \email j.trimble.1@research.gla.ac.uk \\
       \addr University of Glasgow, Glasgow, Scotland}
\maketitle

@@

@@

\section{Non-induced Subgraph Isomorphisms}\label{section:non-induced}

@@

The line plots arithmetic mean search effort using the Glasgow solver: for sparse patterns, the problem is
trivial; for dense patterns, proving unsatisfiability is not particularly difficult; and we see a
complexity peak around the point where half the instances are satisfiable.  We also plot the search
cost of individual instances, as points. The behaviour we observe looks remarkably similar to random
3SAT problems---compare, for example, Figure~1 of the work of
\citeA{DBLP:journals/cacm/Leyton-BrownHHX14}. In particular, satisfiable instances tend to be
easier, but show greater variation than unsatisfiable instances, and there are exceptionally hard
satisfiable instances \cite{DBLP:conf/cp/SmithG97}.

@@

For larger patterns, more complex behaviour emerges. If we are in the intersection of the bottom half
and the bottom right diagonal of the search space, we should always retain the original heuristic,
and if we are in the intersection of the top half and the top left diagonal, we should always use
the complements. This behaviour can be predicted by taking the partial derivatives of $\langle Sol
\rangle$ in the $-d_p$ and $d_t$ directions.  However, when inside the remaining two eighths of the
parameter space, the partial derivatives of $\langle Sol \rangle$ disagree on which heuristic to
use, and using directional derivatives is not enough to resolve the problem. A close observation of
the data suggests that the actual location of the phase transition may be involved---and perhaps
\citeS{DBLP:conf/aaai/Walsh98} suggestion applies only in these conditions. In any case, $\langle Sol
\rangle$ is insufficient to explain the observed behaviour in these two eighths of the parameter space.

@@

The \emph{association graph encoding} of a subgraph isomorphism problem is a reduction to the clique
decision problem. \citeA{DBLP:conf/cp/McCreeshNPS16} describe and study this approach in more
detail. Briefly, the association graph is constructed by creating a new graph with a vertex for each
pair $(p, t)$ of vertices from the pattern and target graphs respectively. There is an edge between
vertex $(p_1, t_1)$ and vertex $(p_2, t_2)$ if mapping $p_1$ to $t_1$ and $p_2$ to $t_2$
simultaneously is permitted, i.e.\ $p_1$ is adjacent to $p_2$ if and only if $t_1$ is adjacent to
$t_2$. A clique of size equal to the order of the pattern graph exists in the association graph if
and only if the problem is satisfiable \cite{o:Levi73}. We used this encoding with an
implementation of \citeS{DBLP:journals/cor/SegundoRJ11} bit-parallel maximum clique algorithm
BBMC, modified to solve the decision problem rather than the optimisation problem.  Again, our
results show that the instances in the central region remain hard, and additionally, some of the
easy unsatisfiable instances become hard.

@@

These experiments further highlight that VF2 occasionally finds some instances which should be easy hard.
But does this cause problems in practice? The literature suggests that it does. For example,
\citeA{o:Gromping14} uses VF2 in a package for the R statistics language, and states
\begin{displayquote}``There are (not so many) instances for which creation of a clear design is
    prohibitively slow in the current implementation that evaluates subgraph isomorphism with the
    VF2 algorithm \ldots Recent experiences with a few of these showed that the LAD algorithm was
very fast in ruling out impossible matches, where VF2 took a long time.'' (p.\ 44)\end{displayquote}
Similarly, \citeA{o:Murray12} uses VF2 inside a compiler, and observes extremely variable
(and prohibitively high) compile times in some cases due to the expense of subgraph isomorphism
calls---given the heavily labelled nature of these graphs, we conjecture that any domain-based
algorithm would eliminate this cost.

@@

Subsequently, in a widely cited\footnote{685 citations according to Google Scholar in February
2018.} paper, \citeA{DBLP:conf/sigmod/YanYH04} introduce an indexing system
called gIndex. Again, this system handles queries by first producing a set of candidates by
eliminating certain unsatisfiable instances, this time by using substructures. They argue that the
query response time, which is to be minimised, is governed by the equation \[ T =
T_{\mathit{search}} + \left|C_q\right| \cdot T_{\mathit{iso\_test}} \textnormal{,}\] where
$T_{\mathit{search}}$ is the time taken to search for a candidate set of potential solutions of size
$\left|C_q\right|$, and $T_{\mathit{iso\_test}}$ is the cost of a subgraph isomorphism test. They
reason that since subisomorphism testing is \NP-complete, by making $\left|C_q\right|$ as small as
possible, the query response time will be reduced.  (Importantly, it is \emph{not} the time taken to
load graphs from disk which contributes to the per-candidate cost, but rather the time to perform a
subgraph isomorphism call.) They conclude that ``graph indexing plays a critical role at efficient
query processing in graph databases'' (p.\ 345).

This equation is repeated and expanded upon by \citeA{DBLP:journals/tods/YanYH05} to explicitly
separate I/O and subisomorphism testing costs into $T_{\mathit{io}}$ and $T_{\mathit{iso\_test}}$
respectively, obtaining a query response time of \[
    T_{\mathit{search}} + \left|C_q\right| \cdot (T_{\mathit{io}} + T_{\mathit{iso\_test}})
    \textnormal{.} \] The authors explicitly state that ``the value of $T_{\mathit{iso\_test}}$ does
not change much for a given query'' (p.\ 996). The argument presented is that
\begin{displayquote}``Sequential scan is very costly
because one has to not only access the whole graph database but also check subgraph isomorphism. It
is known that subgraph isomorphism is an \NP-complete problem. Clearly, it is necessary to build
graph indices in order to help processing graph queries.'' (p.\ 961)\end{displayquote}
We reassess these claims in light of
what we now know about the behaviour of modern subgraph isomorphism algorithms and the nature
of solving \NP-complete problems. Contrary to the authors' claims,
we do not expect $T_{\mathit{iso\_test}}$ to be anything like a constant, even if the orders of the
input graphs are similar. In particular, any instance which can be excluded based upon filtering
must have a very small proof of unsatisfiability. These instances \emph{should} be trivial with any
domain-based subgraph isomorphism algorithm. Thus, all filtering \emph{should} be doing is eliminating the startup
costs of a trivial subgraph isomorphism call. The fact that filtering was successful empirically
should make us wonder whether the subgraph isomorphism algorithm used was substantially
weaker than the state of the art.
Indeed, the algorithm is described only as ``the simplest approach'' in the
paper.  Additionally, the experiments focus on reducing the size of the candidate set, without
considering the time taken to verify different candidate set instances.  The claim that
$T_{\mathit{iso\_test}}$ does not change much is not justified experimentally, and no consideration
is given as to whether this would hold true for other subgraph isomorphism algorithms.

@@

The equation is also used by \citeA{DBLP:conf/icde/JiangWYZ07}, who claim that ``usually the
verification time dominates the Query Response Time [s]ince the computational complexity of
$T_{\mathit{iso\_test}}$ is \NP-Complete''. They note that \begin{displayquote}``Approximately, the
    value of $T_{\mathit{iso\_test}}$ does not change too much with the difference of query. Thus,
the key to reducing query response time is to minimize the size of the candidate answer
set''.  (p.\ 568)\end{displayquote} This claim becomes understandable when one examines the subgraph
isomorphism algorithm chosen for the verification step \cite{DBLP:journals/jacm/Ullmann76}: as the
algorithm predates techniques like generalised arc-consistent all-different propagation
\cite{DBLP:conf/aaai/Regin94}, it cannot immediately detect unsatisfiability in simple cases like
there being two red vertices in the pattern but only one in the target.\footnote{Although
interestingly, this algorithm effectively maintains (binary) arc consistency, and has a
variable-ordering heuristic, before these concepts appeared in the constraints literature.}

Without using the equation, \citeA{DBLP:conf/sigmod/ChengKNL07} state that since subgraph
isomorphism is \NP-complete, processing by a sequential scan is infeasible. They introduce new
filtering techniques to try to avoid the subgraph isomorphism step. A similar claim is made by
\citeA{DBLP:conf/icde/ZhangHY07} in an introduction of another indexing technique: ``obviously it is
inefficient to perform a sequential scan on every graph in the database, because the subgraph
isomorphism test is expensive'' (p.\ 966).

Muddying the waters slightly, a survey by \citeA{DBLP:journals/datamine/HanCXY07} states that
``large volumes of data'' (not \NP-completeness) is the reason for using indexing in these systems.
However, in a description of a system tailored to biological networks,
\citeA{DBLP:conf/edbt/ZhangLY09} state that \begin{displayquote}``Since the size of the raw database
    graph is small, it can be easily fit in the main memory.  However, the query (matching) time
will be very long due to the \NP-hard complexity.'' (p.\ 193)\end{displayquote} They suggest that
indexing is a way of avoiding this. Similarly, \citeA{DBLP:journals/pvldb/ZhaoH10} argue that ``the
graph query problem is hard in that \ldots it requires subgraph isomorphism checking \ldots which
has proven to be \NP-complete'' and that working with large networks is hard or impossible ``due to
the lack of scalable graph indexing mechanisms'' (p.\ 340).

\citeA{DBLP:conf/icdcs/CaoYWRL11} propose a privacy-preserving cloud graph database system using
filter~/ verify, stating that ``checking subgraph isomorphism is \NP-complete, and therefore it is
infeasible to employ such a costly solution'' (p.\ 393) which simply checks every graph for a match.
\citeA{DBLP:journals/tkde/WangWYY12} look at indexing large sparse graphs. They state that ``because
subgraph isomorphism is an \NP-complete problem, a filter-and-verification method is usually
employed to speed up the search efficiency of graph similarity matching over a graph set'' (p.\ 441).
Similarly, after reviewing the literature, \citeA{DBLP:journals/vldb/YuanM13} argue that subgraph
querying is costly because it is \NP-complete, and that indices can improve the performance of graph
database queries. Again, new indexing techniques are introduced. Subsequently,
\citeA{DBLP:journals/pvldb/KatsarouNT15} perform a comprehensive comparison of graph database
filtering techniques. They state that performing a query against each graph in the dataset
``obviously does not scale, as subgraph isomorphism is \NP-complete'' (p.\ 1566). In describing a
system which returns a special subset of graphs which match a query,
\citeA{DBLP:journals/tkde/ZhengLZHZ16} suggest that it is ``\NP-hard to check the graph
isomorphism'' (meaning subgraph isomorphism), and ``in order to improve the time efficiency'' they
use an indexing system to ``avoid as many costly subgraph isomorphism checkings as possible'' (p.\
1806). Their index takes tens of thousands of seconds to build, and they suggest that Ullmann's
algorithm and VF2 are state of the art for verification.  \citeA{DBLP:journals/www/PengZ0LZ16} state
that ``obviously, it is impossible to employ some subgraph isomorphism algorithm, such as Ullmann or
VF2'', and argue that ``in order to speed up query processing'', they need to create indices (p.\
418).  And \citeA{DBLP:conf/isda/AzaouziR16} argue that ``since the subgraph isomorphism test is expensive,
checking all graphs of a large database can be unfeasible'', saying that ``naturally, the
verification step is computationally more expensive since it requires a subgraph isomorphism'' (p.\
251); their experiments look at candidate set reduction sizes, and their choice of subgraph isomorphism
algorithm is not mentioned.

The filter~/ verify paradigm also influences other research. For example, it is used by
\citeA{DBLP:conf/icde/TianP08} in an approximate subgraph searching system: they state that
Ullmann's algorithm ``is prohibitively expensive for querying against [a] database with a large
number of graphs'', and that indices are used ``to filter out graphs that do not match the query''
(p.\ 964).  More recently, \citeA{DBLP:journals/pvldb/YuanMG13} continue a line of supergraph search work, again
using a filtering step to avoid subgraph isomorphism calls (in the opposite direction, so queries
are now target graphs). \citeA{DBLP:journals/tkde/HongZLY15} look at graph database subgraph
isomorphism with an additional set-similarity constraint, and state that Ullmann's algorithm and VF2
are ``usually costly for large graphs'' because they ``do not utilize any index structure'' (p.\
2509). They propose an indexing structure, which takes over 2,000 seconds to construct, and uses nearly 2GBytes
of space.  There is also research into maintaining indices when the set of target graph changes: for
example \citeA{DBLP:conf/sigmod/YuanMYG15} look at algorithms for updating graph indices.
And describing a system for reusing results of queries which are sub- or super-graphs of previous
queries, \citeA{DBLP:conf/edbt/WangNT16} state that querying is a ``very costly operation as it
entails the \NP-complete problem of subgraph isomorphism'' (p.\ 41), and place ``an emphasis on the
number of unnecessary subgraph isomorphism tests'' (p.\ 42).

@@

We saw in the previous sections that although permuting input graphs could improve VF2's behaviour
on some instances, doing so does not make its performance come close to that of domain-based
algorithms.  \citeA{DBLP:conf/edbt/KatsarouNT17} say they ``hope that our findings will open up
new research directions, striving to find appropriate, per-query, isomorphic rewritings, in
combination with alternate per-query sub-iso algorithms that can yield large improvements''
(p.\ 36). We believe it is important instead to emphasise previous research directions that have already
solved most of this problem, and to help ensure that these techniques become more widely known.

\citeS{DBLP:conf/edbt/KatsarouNT17} second claim is that ``different
algorithms find different queries hard'' (p.\ 30). To address this, they run many subgraph isomorphism
algorithms and input permutations in parallel; the extensive literature on parallel portfolios in
general \cite{DBLP:journals/ai/GomesS01}, and the approach by \citeA{DBLP:conf/sls/BattitiM07} for
subgraph isomorphism in particular, is not noted. (Nor is the portfolios literature referenced when
they say that ``using machine learning models to predict which version \ldots to employ per query is
of high interest'', p.\ 36).  The evidence so far in this paper suggests that we should carefully examine
the reasons behind these speedups. It is certainly true that there are some instances that all
algorithms find hard, and there are good theoretical reasons to believe that these instances
genuinely are really hard.  Furthermore, algorithm portfolios are also well-known to be a successful
technique, and are beneficial even with modern subgraph isomorphism algorithms
\cite{DBLP:conf/lion/KotthoffMS16}.  However, in \cref{section:labelled} we saw that there are many
instances that VF2 finds hard that should not be hard, and that are not hard for other algorithms.
If permuting graphs sometimes addresses the difficulty of some of these instances, then it is likely
that they are not genuinely hard instances at all. Perhaps then, it would be better simply to use an
algorithm with domain tracking and domain-based ordering heuristics.

To test this suggestion, we return briefly to the experiments on the datasets discussed earlier in
this section. What if we modify our solver so that it does not use ``smallest domain first'', and
does not detect domain wipeouts until attempting to branch on an empty variable?  In this case, three of
the four datasets become extremely hard, with many instances now timing out after making hundreds of
millions of recursive calls, and the PPI dataset now requires hundreds of thousands of calls for
some queries.  Furthermore, slight changes to the input vertex ordering can now make some of these
``hard'' instances easy again.  This verifies our intuition: the apparent success of
\citeS{DBLP:conf/edbt/KatsarouNT17} technique is due to the ease of sometimes getting better
results out of a poor algorithm. Obtaining huge improvements from an algorithm portfolio consisting
only of variations of poor algorithms is not sufficient to demonstrate a genuine improvement over
the state of the art.

@@

\begin{displayquote}
    ``Benchmark problems should be hard. I report on the solution of the five open benchmark
    problems introduced \ldots for testing bin packing problems. Since the solutions were found
    either by hand or by using very simple heuristic methods, these problems would appear to be
    easy. In four cases I give improved packings to refute conjectures that previously reported
    packings were optimal, and I give a proof that the fifth conjecture was correct. \ldots Future
    experimenters should be careful to perform tests on problems that can reasonably be regarded as
    hard.'' (p.\ 299)
\end{displayquote}

@@

\citeA{o:CarlettiFSV17} correctly observe that ``a performance improvement on [hard instances] has a
far greater practical impact than it would have on easier graphs, since the saved time can be of
several hours or even days'' (p.\ 805). To evaluate VF3, they introduce a new dataset consisting
entirely of satisfiable instances created by taking a connected subgraph of a randomly generated
graph and permuting the nodes. They conclude that VF3 is the best algorithm based upon its
performance solving the enumeration problem on these new instances.  We urge a more cautious
interpretation of these results: as we saw in \cref{section:induced}, being genuinely hard (rather
than simply larger or denser) is a subtle property, and care must be taken when empirically
evaluating algorithms. In particular, if one only considers large, easy instances, then the simpler
the algorithm, the better it scales. Indeed, we saw in \cref{figure:induced} that VF3 performs
\emph{worse} than VF2 on genuinely harder instances, and additionally behaves erratically in parts
of the parameter space that every other solver finds easy; finally, an induced analogue of
\cref{figure:labels} confirms that VF3 also still struggles with many obviously unsatisfiable
instances.

@@

\acks{
The authors would like to thank Kitty Meeks and Craig Reilly for their comments. We are grateful to
Iva Babukova for her help with the datasets and for sharing the results of her experiments.  The
heatmap schemes used in this paper were created with the assistance of the ``Chroma.js Color Scale
Helper'' by Gregor Aisch.  Parts of this paper were presented at IJCAI
\cite{DBLP:conf/ijcai/McCreeshPT16}; the first three sections have been expanded, whilst
\cref{section:regular,section:labelled,section:filterverify} are new material. Most of this work
also appeared in a PhD thesis \cite{o:McCreesh17}.

This work was supported by the Engineering and Physical Sciences Research Council [grant numbers
EP/K503058/1, EP/M508056/1, and EP/P026842/1], and by the ANR project SoLStiCe
(ANR-13-BS02-0002-01).}
